Individual and Differential Harm in Redistricting

### 2022-08-18 paper/harm.Rmd
further tweaks

To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^12] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
The turnout component of the model is identical to that used for Alabama.
The core of the voter choice model is the following specification for $\pi_{jk}$, the preference of precinct $j$ for candidate $k$ on the logit scale:[^13] \begin{align*}
    \pi_{jk} &= \underbrace{\beta^{(s)}_k}_{\substack{\text{citywide support}\\\text{for candidate}}} +
    \pref_{j}^\top(\overbrace{P_j}^{\mathclap{\substack{\text{diagonal matrix of}\\\text{precinct racial proportions}}}}
    \underbrace{\turn_j}_{\mathclap{\substack{\text{turnout rate parameter}\\\text{from turnout model}}}}
    )^\top\vb L_k.
\end{align*} Here, $\pref_j$ is a matrix of preferences by race and latent dimension within precinct $j$, so $\pref_{j}^\top(P_j\turn_j)^\top$ is a vector with elements containing the preferences of the precinct's voters in each latent dimension, after marginalizing out race.
The vector $\vb L_k$ contains the loading of candidate $k$ onto the latent space: When $\vb L_k$ and $\pref_{j}$ are nearby in latent space, then their product will be large, and the precinct's support for the candidate, $\pi_{jk}$, will be large.
Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R. Biden and Wu opponent Anissa Essaibi-George on the right.
The second dimension appears primarily to capture racial differences in support---as the factor value maps in Figure \ref{fig:boston-factors} show, minority-majority and White areas of the city score opposite values on the second dimension.

@@

Continuing the notation from above, for precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $n_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jk}$ denote the vote share for candidate $k$ (who implicitly belongs to a particular election $v_k$), and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Hispanic},\text{Other}\}$ ("Other" consisting primarily of Asian voters in Boston).
Our main parameters of interest are $\turn_{jr}$, the baseline turnout by race and precinct, $\pref_{jqr}$, the preferences in latent dimension $q$ by race and precinct, and $L_{kq}$, the loading of each candidate on latent dimension $q$.
Here $q\in\{1,2\}$, since preliminary analyses found no improved fit when expanding to three dimensions.

@@

The logit candidate preference itself is modeled as \begin{align*}
    \pi_{jk} &= \beta^{(s)}_k + \alpha_k
        \pref_{j}^\top(\turn_j\circ \vb p_j)^\top\mathrm{diag}(\vb*\rho)\vb L_k  \\
    \pref_{jq} \mid \Sigma^{(p)} &\sim \Norm(0, \Sigma^{(p)}).
\end{align*} Note that $\pref_j$ is a $q\times r$ matrix of preferences by race and dimension within precinct $j$, and $\turn_j\circ \vb p_j$ is a vector of the percentages of the total VAP in the precinct that belongs to voters of each race, so $\pref_{j}^\top(\turn_j\circ \vb p_j)$ is a vector with $q$ elements containing the preferences by dimension of the precinct's voters, after marginalizing out race.
The matrix $\mathrm{diag}(\vb*\rho)$ captures the relative scale or importance of each dimension to the overall outcomes.
The $\vb*\alpha$ parameters are "fudge factors" that adjust the scale of the deviations from each candidate's average support.

@@

The estimated loading matrix $\vb L$ is visualized on the left of Figure \ref{fig:boston-factors}, and $\pref$ is visualized for Black and White voters on the right.

>>>>

To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^12] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
The turnout component of the model is identical to that used for Alabama.
The core of the voter choice model is the following specification for $\pi_{jk}$, the preference of precinct $j$ for candidate $k$ on the logit scale:[^13] \begin{align*}
    \pi_{jk} &= \underbrace{\beta^{(s)}_k}_{\substack{\text{citywide support}\\\text{for candidate}}} +
    F_{j}^\top(\overbrace{P_j}^{\mathclap{\substack{\text{diagonal matrix of}\\\text{precinct racial proportions}}}}
    \underbrace{\turn_j}_{\mathclap{\substack{\text{turnout rate parameter}\\\text{from turnout model}}}}
    )^\top\vb L_k.
\end{align*} Here, $F_j$ is a matrix of preferences by race and latent dimension within precinct $j$, so $F_{j}^\top(P_j\turn_j)^\top$ is a vector with elements containing the preferences of the precinct's voters in each latent dimension, after marginalizing out race.
The vector $\vb L_k$ contains the loading of candidate $k$ onto the latent space: When $\vb L_k$ and $F_{j}$ are nearby in latent space, then their product will be large, and the precinct's support for the candidate, $\pi_{jk}$, will be large.
Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R. Biden and Wu opponent Anissa Essaibi-George on the right.
The second dimension appears primarily to capture racial differences in support---as the factor value maps in Figure \ref{fig:boston-factors} show, minority-majority and White areas of the city take opposite values on the second dimension.

@@

Continuing the notation from above, for precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $n_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jk}$ denote the vote share for candidate $k$ (who implicitly belongs to a particular election $v_k$), and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Hispanic},\text{Other}\}$ ("Other" consisting primarily of Asian voters in Boston).
Our main parameters of interest are $\turn_{jr}$, the baseline turnout by race and precinct, $F_{jqr}$, the preferences in latent dimension $q$ by race and precinct, and $L_{kq}$, the loading of each candidate on latent dimension $q$.
Here $q\in\{1,2\}$, since preliminary analyses found no improved fit when expanding to three dimensions.

@@

The logit candidate preference itself is modeled as \begin{align*}
    \pi_{jk} &= \beta^{(s)}_k + \alpha_k
        F_{j}^\top(\turn_j\circ \vb p_j)^\top\mathrm{diag}(\vb*\rho)\vb L_k  \\
    F_{jq} \mid \Sigma^{(p)} &\sim \Norm(0, \Sigma^{(p)}).
\end{align*} Note that $F_j$ is a $q\times r$ matrix of preferences by race and dimension within precinct $j$, and $\turn_j\circ \vb p_j$ is a vector of the percentages of the total VAP in the precinct that belongs to voters of each race, so $F_{j}^\top(\turn_j\circ \vb p_j)$ is a vector with $q$ elements containing the preferences by dimension of the precinct's voters, after marginalizing out race.
The matrix $\mathrm{diag}(\vb*\rho)$ captures the relative scale or importance of each dimension to the overall outcomes.
The $\vb*\alpha$ parameters are "fudge factors" that adjust the scale of the deviations from each candidate's average support.

@@

The estimated loading matrix $\vb L$ is visualized on the left of Figure \ref{fig:boston-factors}, and $F$ is visualized for Black and White voters on the right.
